---
title: "Auto Level Selection with auto_segment_levels()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{monecascale-auto-level}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5,
  dpi = 150,
  out.width = "100%"
)

# Eval-gate on greed (Suggests). If missing, skip evaluation and still
# render a readable article.
greed_available <- requireNamespace("greed", quietly = TRUE)
if (!greed_available) {
  knitr::opts_chunk$set(eval = FALSE)
}

ggplot2_available <- requireNamespace("ggplot2", quietly = TRUE)
```

## 1. Introduction

`auto_segment_levels()` closes a recurring loose end in hierarchical
segmentation: the backend returns a full nested hierarchy, but downstream
analysis typically needs a single level. Picking that level by hand is
brittle, does not scale, and is difficult to justify in a reproducible
pipeline.

The function implements two principled, purely post-hoc criteria for
picking a preferred level from a fitted hierarchy. It composes directly
with the two scalable backends in this package:

- `moneca_sbm()` (direction D2) - hierarchical DC-SBM on square mobility.
- `moneca_bipartite()` (direction D1) - hierarchical DC-LBM on
  rectangular mobility, picked independently on the row and column
  sides.

Plain `moneca::moneca_fast()` output is not yet supported and is
rejected with an informative error; that hook is deferred to
monecascale 0.3.1.

Two entry points exist. The function itself takes a fitted object and
returns the pick plus per-level diagnostics. The `segment.levels =
"auto"` argument on `moneca_sbm()` and `moneca_bipartite()` is the
wrapper sugar: fit the full hierarchy, pick a level, and trim the
returned object to that level in one call.

## 2. Criteria

**MDL elbow.** The criterion walks the per-level Minimum Description
Length (MDL) series (coarse to fine) and picks the point of maximum
perpendicular distance from the chord connecting the two endpoints.
This is the Kneedle heuristic (Satopaa et al. 2011) and is the default.
An alternative `mdl_elbow = "max_second_diff"` selects the discrete
second-difference peak, which is more sensitive on short hierarchies
but noisier on long ones. On `moneca_sbm()` the MDL series is `-ICL`
from `greed::DcSbm`; on `moneca_bipartite()` the per-side series is
`-joint_icl` from `greed::DcLbm`, sliced to the side at hand. In both
cases lower is better, so Kneedle detects the knee of a decreasing
curve.

**MI plateau.** The criterion computes level-to-level mutual
information `I(m_l, m_{l+1})` between adjacent hierarchy levels
directly from `segment.list`, independently of backend diagnostics.
`I(m_l, m_{l+1})` measures how much structure level `l+1` shares with
the coarser level `l`: it is high when refinement is consistent with
what was there at the coarser level and drops where refinement begins
to reshuffle memberships. The plateau is the first level whose gain
`I_l - I_{l-1}` falls below `plateau_tol * max(I)` - the resolution at
which further refinement adds little and is therefore heuristically
redundant. If no plateau is detected the finest level is returned.

The two criteria use different inputs (MDL series vs. membership
vectors), so they can and often do disagree. That disagreement is
itself diagnostic, as illustrated in section 6.

## 3. Quick start on SBM

The quickest way to see the function at work is on a synthetic
mobility fixture small enough to build in a vignette. We keep the
fixture inline rather than sourcing the test helpers.

```{r sbm-quick-fit}
library(monecascale)

# 1. synthetic fixture -----
mob <- moneca::generate_mobility_data(
  n_classes = 20,
  n_total = 5000,
  immobility_strength = 0.5,
  class_clustering = 0.75,
  noise_level = 0.05,
  seed = 2026
)

# 2. full hierarchy fit -----
fit <- monecascale::moneca_sbm(mob, backend = "greed", seed = 2026)

# 3. pick a level post-hoc -----
res <- monecascale::auto_segment_levels(fit, method = "mdl")
res
```

The printed header exposes method, backend, picked level, and the
total number of levels available. The diagnostics table holds the
full per-level trace used to make the pick, which is useful both as
an audit record and for plotting the criterion curve.

```{r sbm-diag}
res$diagnostics
```

Switching criterion is a one-argument change:

```{r sbm-mi}
res_mi <- monecascale::auto_segment_levels(fit, method = "mi_plateau")
res_mi
```

`method = "mi_plateau"` ignores `mdl_per_level` and operates directly
on the cliques stored in `segment.list`, so it works unchanged on any
hierarchy produced by this package.

## 4. Wrapper sugar: segment.levels = "auto"

`moneca_sbm()` and `moneca_bipartite()` accept `segment.levels =
"auto"`. The wrapper fits the full hierarchy, delegates to
`auto_segment_levels()`, and trims the returned object down to the
picked level. The picker result is attached so callers can still
inspect what happened.

```{r sbm-auto-wrapper}
fit_auto <- monecascale::moneca_sbm(
  mob,
  backend = "greed",
  seed = 2026,
  segment.levels = "auto",
  auto_method = "mdl"
)

# The full-hierarchy fit would have returned length(segment.list) levels;
# after trimming we see only those up to the pick.
length(fit_auto$segment.list)

# The picker result is retained for auditability.
fit_auto$auto_level

# The object is otherwise a plain moneca-class object.
class(fit_auto)
```

The trimmed object is a regular moneca-class object: the
`moneca::segment.*()` and `moneca::plot_moneca_*()` APIs run on it
unchanged. `auto_method = "mi_plateau"` is accepted symmetrically and
dispatches to the MI criterion inside the wrapper.

## 5. Bipartite

On a bipartite fit the two sides are clustered jointly but exposed
through two one-mode projections. `auto_segment_levels()` recurses
into `$rows` and `$cols`, runs the criterion independently on each
side, and returns a named list. The two sides may land on different
levels.

```{r bip-fixture}
# 1. rectangular fixture with planted row / column blocks -----
set.seed(2026)
n_row <- 60
n_col <- 40
row_block <- rep(seq_len(4), length.out = n_row)
col_block <- rep(seq_len(3), length.out = n_col)
lambda <- outer(row_block, col_block, function(g, h) {
  ifelse((g - 1) %% 3 + 1 == h, 6, 1)
})
mx <- matrix(stats::rpois(n_row * n_col, lambda = lambda), n_row, n_col)
rownames(mx) <- paste0("R", seq_len(n_row))
colnames(mx) <- paste0("C", seq_len(n_col))

# 2. full bipartite fit -----
bip_fit <- monecascale::moneca_bipartite(mx, seed = 2026, verbose = FALSE)
```

```{r bip-pick}
bip_res <- monecascale::auto_segment_levels(bip_fit, method = "mdl")

# Two independent picks, one per side.
names(bip_res)
bip_res$rows
bip_res$cols
```

The wrapper form behaves the same way as on SBM, except that the
trimming is applied per side:

```{r bip-auto-wrapper}
bip_auto <- monecascale::moneca_bipartite(
  mx,
  seed = 2026,
  verbose = FALSE,
  segment.levels = "auto",
  auto_method = "mdl"
)

# Per-side picks; the two sides may disagree.
bip_auto$rows$auto_level$level
bip_auto$cols$auto_level$level
length(bip_auto$rows$segment.list)
length(bip_auto$cols$segment.list)
```

## 6. Criterion comparison

MDL elbow and MI plateau see different signals: the first operates on
a backend goodness-of-fit trace, the second on the memberships the
hierarchy actually delivered. On the same fit we can run both and
inspect the two curves side by side.

```{r compare-run}
res_mdl <- monecascale::auto_segment_levels(fit, method = "mdl")
res_mi  <- monecascale::auto_segment_levels(fit, method = "mi_plateau")

# Picks
c(mdl = res_mdl$level, mi_plateau = res_mi$level)
```

The diagnostics tables carry the series each criterion used.

```{r compare-tables}
res_mdl$diagnostics[, c("level", "n_blocks", "mdl")]
res_mi$diagnostics[, c("level", "n_blocks", "mi_to_next")]
```

If `ggplot2` is available we can plot the two curves with the picks
annotated. Otherwise the tables above are the portable summary.

```{r compare-plot, eval = greed_available && ggplot2_available, fig.height = 7}
# 1. prepare panels -----
mdl_df <- res_mdl$diagnostics
mdl_df$criterion <- "MDL (lower is better)"
mdl_df$value <- mdl_df$mdl

mi_df <- res_mi$diagnostics
mi_df$criterion <- "MI to next level"
mi_df$value <- mi_df$mi_to_next

plot_df <- rbind(
  mdl_df[, c("level", "n_blocks", "value", "criterion")],
  mi_df[, c("level", "n_blocks", "value", "criterion")]
)
plot_df <- plot_df[is.finite(plot_df$value), ]

# 2. pick annotations -----
picks <- data.frame(
  criterion = c("MDL (lower is better)", "MI to next level"),
  level = c(res_mdl$level, res_mi$level)
)
picks$value <- mapply(
  function(cr, lvl) {
    sub <- plot_df[plot_df$criterion == cr & plot_df$level == lvl, "value"]
    if (length(sub) == 0) NA_real_ else sub[1]
  },
  picks$criterion,
  picks$level
)

# 3. stacked panel -----
ggplot2::ggplot(
  plot_df,
  ggplot2::aes(x = level, y = value)
) +
  ggplot2::geom_line() +
  ggplot2::geom_point() +
  ggplot2::geom_vline(
    data = picks,
    mapping = ggplot2::aes(xintercept = level),
    linetype = 2
  ) +
  ggplot2::geom_point(
    data = picks[is.finite(picks$value), ],
    mapping = ggplot2::aes(x = level, y = value),
    colour = "#0054AD",
    size = 3
  ) +
  ggplot2::facet_wrap(~ criterion, ncol = 1, scales = "free_y") +
  ggplot2::labs(
    x = "segment.list level",
    y = NULL,
    title = "MDL elbow vs. MI plateau on the same SBM fit"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

The two criteria frequently agree on well-structured data; when they
disagree, MDL tends to pick a coarser level than MI plateau on
hierarchies where refinement keeps consistent sub-blocks but the MDL
curve is still trending down.

## 7. Limitations

- **`moneca::moneca_fast()` output is not supported.** The function
  errors out on fast-backend objects. Deferred to 0.3.1.
- **Single-criterion set.** Only MDL elbow and MI plateau are shipped
  in 0.3.0. Modularity-based picks and bootstrap stability criteria
  are on the roadmap but not in scope for this release.
- **`plateau_tol` is a heuristic.** The default 0.05 works well on
  moderate hierarchies but may need tightening on tight elbows or
  loosening on noisy MI curves. The argument is surfaced on
  `auto_segment_levels()` for that reason.
- **Short hierarchies degrade gracefully.** On fits with at most two
  levels in `segment.list`, both criteria fall back to the finest
  level. The diagnostics table still reports per-level values where
  they can be computed.

## 8. References

- Satopaa, V., Albrecht, J., Irwin, D., & Raghavan, B. (2011).
  Finding a "Kneedle" in a Haystack: Detecting Knee Points in System
  Behavior. *31st International Conference on Distributed Computing
  Systems Workshops*, 166-171.
- Cover, T. M., & Thomas, J. A. (2006). *Elements of Information
  Theory* (2nd ed.). Wiley.
- Come, E., Jouvin, N., Latouche, P., Bouveyron, C., Bataillon, E., &
  Chen, A. (2021). Hierarchical clustering with discrete latent
  variable models and the integrated classification likelihood.
  *Advances in Data Analysis and Classification*, 15, 957-986.
- Karrer, B., & Newman, M. E. J. (2011). Stochastic blockmodels and
  community structure in networks. *Physical Review E*, 83, 016107.

```{r no-greed-note, echo = FALSE, eval = !greed_available}
cat(
  "Note: The `greed` package was not available when this vignette was\n",
  "built. The code chunks above were not evaluated. Install `greed`\n",
  "with install.packages('greed') to reproduce the examples.\n",
  sep = ""
)
```
