---
title: "Bipartite Segmentation with moneca_bipartite()"
author: "Giampaolo Montaletti"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{monecascale-bipartite}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5,
  dpi = 150,
  out.width = "100%"
)

# Eval-gate on greed (Suggests). If missing, skip evaluation and still
# render a readable article.
greed_available <- requireNamespace("greed", quietly = TRUE)
if (!greed_available) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

## 1. Introduction

`moneca_bipartite()` extends the MONECA analysis stack to rectangular
(two-mode) mobility data, where rows and columns are distinct entity
types. Typical use cases are person x employer, worker x occupation,
and student x school tables. The square mobility matrix that
`moneca::moneca_fast()` and `monecascale::moneca_sbm()` operate on is
already a one-mode aggregation; bipartite input preserves individual
row and column identity and lets the block model infer groups on both
sides jointly.

This vignette is the rectangular counterpart of
`vignette("monecascale-sbm", package = "monecascale")`: the square
case motivates why relative risk is the DC-SBM residual, this one
motivates why the same argument carries over to the Latent Block Model
(LBM) and how the joint row / column clustering is exposed to moneca's
downstream `segment.*()` and `plot_moneca_*()` APIs.

## 2. Theoretical bridge: rectangular RR as DcLbm residual

MONECA's weight is the relative-risk deviation of observed mobility
from the independence null. For a rectangular core `O` with row
margins `r`, column margins `c`, and grand total `N`:

$$
RR_{ij} = \frac{O_{ij}}{E_{ij}}, \qquad E_{ij} = \frac{r_i \cdot c_j}{N}.
$$

The degree-corrected Latent Block Model (DcLbm, Come et al. 2021)
models the expected edge count between row `i` in row-block `g` and
column `j` in column-block `h` as
`theta_i^row * theta_j^col * omega_{g,h}`, with `theta` degree
parameters and `omega` the row-block by column-block interaction
matrix. Setting `omega` to the all-ones matrix recovers exactly
`(r_i c_j) / N`, i.e. MONECA's independence null. DcLbm inference on
the rectangular mobility is therefore the principled scalable drop-in
for clique enumeration on the bipartite case, in the same way that
DcSbm is for the unipartite case covered in `monecascale-sbm`.

The current backend is `greed::DcLbm`, an R-native hierarchical
implementation built on Integrated Classification Likelihood with
merge-split local search. A `"graphtool"` backend via
`graph_tool.inference.minimize_blockmodel_dl(pclabel = ...)` is
reserved for scale beyond 10^5 per side.

## 3. Quick start

```{r quick-start}
library(monecascale)

# 1. synthetic rectangular fixture -----
set.seed(2026)
n_row <- 60
n_col <- 40
row_block <- rep(1:4, length.out = n_row)
col_block <- rep(1:3, length.out = n_col)
# block-constant Poisson intensities: diagonal-ish structure
lambda <- outer(row_block, col_block, function(g, h) {
  ifelse((g - 1) %% 3 + 1 == h, 6, 1)
})
mx <- matrix(stats::rpois(n_row * n_col, lambda = lambda), n_row, n_col)
rownames(mx) <- paste0("R", seq_len(n_row))
colnames(mx) <- paste0("C", seq_len(n_col))

# 2. fit -----
out <- monecascale::moneca_bipartite(mx, seed = 2026, verbose = FALSE)

# 3. inspect -----
out
summary(out)
out$bipartite_diagnostics$n_blocks_row_per_level
out$bipartite_diagnostics$n_blocks_col_per_level
```

The `print.moneca_bipartite()` method shows the per-level table of row
block count, column block count, and joint ICL. `summary()` adds the
backend metadata block (`n_init`, `rr_scale`, margin status).

## 4. Output anatomy

`moneca_bipartite()` returns an object of class `"moneca_bipartite"`
(disjoint from `"moneca"`) with three slots: `$rows`, `$cols`, and
`$bipartite_diagnostics`. The first two are full moneca-class objects
built from one-mode projections of the rectangular RR. The third
retains the joint bipartite information that a single square object
cannot represent.

```{r classes}
class(out)
class(out$rows)
class(out$cols)
```

The row view is a square, symmetric, zero-diagonal matrix on `n_row`
entities; the column view is square on `n_col`. Margins are appended
as the last row and column, matching moneca's convention.

```{r shapes}
dim(out$rows$mat.list[[1]])
dim(out$cols$mat.list[[1]])
isSymmetric(as.matrix(out$rows$mat.list[[1]]))
```

The diagnostics slot exposes the joint structure:

```{r diagnostics}
str(out$bipartite_diagnostics, max.level = 1)
out$bipartite_diagnostics$n_blocks_row_per_level
out$bipartite_diagnostics$n_blocks_col_per_level
round(out$bipartite_diagnostics$joint_icl_per_level, 2)
```

`block_interaction_matrix_per_level` is a list of `Krow x Kcol`
matrices, one per hierarchy level, cross-tabulating the rectangular
core by row and column memberships. Level 1 is finest.

```{r joint-matrix}
bim <- out$bipartite_diagnostics$block_interaction_matrix_per_level
length(bim)
dim(bim[[1]])
round(bim[[1]], 1)
```

## 5. Downstream compatibility

Because `$rows` and `$cols` are full moneca-class objects, the entire
moneca consumer API runs on them unchanged. The two views are queried
independently: the row view answers questions about how row entities
cluster given their column-mediated co-movement, and symmetrically for
the column view.

```{r downstream-membership}
head(moneca::segment.membership(out$rows))
head(moneca::segment.membership(out$cols))
```

```{r downstream-quality}
moneca::segment.quality(out$rows)
moneca::segment.quality(out$cols)
```

```{r downstream-ggraph-rows, fig.alt = "Row-view network plot"}
if (requireNamespace("ggraph", quietly = TRUE)) {
  moneca::plot_moneca_ggraph(
    out$rows,
    level = 2,
    node_color = "segment",
    title = "moneca_bipartite - row view"
  )
}
```

```{r downstream-hier-cols, fig.alt = "Column-view hierarchical plot"}
if (requireNamespace("ggraph", quietly = TRUE)) {
  moneca::plot_moneca_hierarchical(
    out$cols,
    title = "moneca_bipartite - column view"
  )
}
```

The joint row-block by column-block view is not a moneca-class object
and is plotted directly from the diagnostics slot:

```{r joint-heatmap, fig.alt = "Joint block-interaction heatmap"}
image(
  t(bim[[1]])[, nrow(bim[[1]]):1],
  axes = FALSE,
  main = "Joint block interaction (finest level)",
  xlab = "column block",
  ylab = "row block"
)
```

## 6. Latent-block recovery experiment

A block-constant generator with known row and column memberships lets
us measure how well DcLbm recovers the planted structure. We compare
the row-side and column-side memberships at the level closest to the
planted sizes against `row_block` and `col_block` with the adjusted
Rand index. The fixture is independent of the quick-start block to
keep this section self-contained.

```{r parity-setup}
# 1. planted design -----
set.seed(42)
n_row <- 80
n_col <- 60
K_row_true <- 4
K_col_true <- 3
row_block_true <- rep(seq_len(K_row_true), length.out = n_row)
col_block_true <- rep(seq_len(K_col_true), length.out = n_col)

# 2. block-constant Poisson with strong diagonal signal -----
omega <- matrix(0.8, K_row_true, K_col_true)
for (g in seq_len(K_row_true)) {
  h <- ((g - 1) %% K_col_true) + 1L
  omega[g, h] <- 8
}
lambda <- omega[row_block_true, col_block_true]
mx_parity <- matrix(
  stats::rpois(n_row * n_col, lambda = lambda),
  n_row,
  n_col
)
rownames(mx_parity) <- paste0("R", seq_len(n_row))
colnames(mx_parity) <- paste0("C", seq_len(n_col))

# 3. fit -----
fit_parity <- monecascale::moneca_bipartite(
  mx_parity,
  seed = 42,
  verbose = FALSE
)
```

```{r parity-helpers}
# Minimal ARI. Returns 1 when labelings agree up to permutation.
ari <- function(a, b) {
  tab <- table(a, b)
  n <- sum(tab)
  sc <- function(x) sum(choose(x, 2))
  a_sum <- sc(rowSums(tab))
  b_sum <- sc(colSums(tab))
  tab_sum <- sc(as.vector(tab))
  expected <- a_sum * b_sum / choose(n, 2)
  max_idx <- (a_sum + b_sum) / 2
  if (max_idx == expected) return(1)
  (tab_sum - expected) / (max_idx - expected)
}

level_labels <- function(mon, level) {
  n_core <- nrow(mon$mat.list[[1]]) - 1L
  labels <- rep(NA_integer_, n_core)
  cliques <- mon$segment.list[[level]]
  for (b in seq_along(cliques)) labels[cliques[[b]]] <- b
  na_idx <- is.na(labels)
  if (any(na_idx)) {
    labels[na_idx] <- max(0L, labels, na.rm = TRUE) + seq_len(sum(na_idx))
  }
  as.integer(labels)
}
```

```{r parity-score}
# 1. pick the level whose row-block count matches the planted K_row -----
krow_per_level <- fit_parity$bipartite_diagnostics$n_blocks_row_per_level
kcol_per_level <- fit_parity$bipartite_diagnostics$n_blocks_col_per_level
row_level <- which.min(abs(krow_per_level - K_row_true)) + 1L
col_level <- which.min(abs(kcol_per_level - K_col_true)) + 1L

# 2. extract moneca-style node labels at those levels -----
row_hat <- level_labels(fit_parity$rows, row_level)
col_hat <- level_labels(fit_parity$cols, col_level)

# 3. score against planted memberships -----
ari_row <- ari(row_block_true, row_hat)
ari_col <- ari(col_block_true, col_hat)
data.frame(
  side = c("rows", "cols"),
  K_true = c(K_row_true, K_col_true),
  K_hat = c(krow_per_level[row_level - 1L], kcol_per_level[col_level - 1L]),
  ARI = round(c(ari_row, ari_col), 3)
)
```

On this fixture DcLbm typically recovers both sides with ARI > 0.5,
and often well above. Exact values vary with the seed and with the
signal-to-noise contrast between diagonal and off-diagonal block
intensities. The load-bearing claim is directional: the rectangular
RR = DcLbm residual bridge is empirically supported on data where the
planted structure is known.

## 7. Limitations and future directions

- **One-mode projections, not joint clustering.** Exposing the fit to
  moneca requires two square objects, so the joint `Krow x Kcol`
  co-clustering is not representable inside `$rows` or `$cols`. The
  full block-interaction matrices are retained in
  `$bipartite_diagnostics$block_interaction_matrix_per_level` for
  consumers that need the joint view.
- **No isolates summary on the bipartite side yet.** Row- and
  column-side isolates surfaces are deferred to release 0.2.1.
- **Single backend.** `moneca_bipartite()` currently exposes the
  `greed` DcLbm backend only. A `graph-tool` bipartite backend via
  `pclabel` is reserved for input sizes beyond ~10^5 per side.
- **Auto-level composition pending.** Until direction **D6**
  (`auto_segment_levels()`) lands, cap the hierarchy explicitly via
  `segment.levels` when downstream consumers expect a fixed depth.
- **Square input is rejected.** For square mobility matrices use
  `monecascale::moneca_sbm()` or `moneca::moneca_fast()`.

## 8. References

- Come, E., Jouvin, N., Latouche, P., Bouveyron, C., Bataillon, E., &
  Chen, A. (2021). Hierarchical clustering with discrete latent
  variable models and the integrated classification likelihood.
  *Advances in Data Analysis and Classification*, 15, 957-986.
- Larremore, D. B., Clauset, A., & Jacobs, A. Z. (2014). Efficiently
  inferring community structure in bipartite networks. *Physical
  Review E*, 90(1), 012805.
- Karrer, B., & Newman, M. E. J. (2011). Stochastic blockmodels and
  community structure in networks. *Physical Review E*, 83, 016107.
- Peixoto, T. P. (2014). Hierarchical block structures and
  high-resolution model selection in large networks. *Physical Review
  X*, 4, 011047.

```{r no-greed-note, echo = FALSE, eval = !greed_available}
cat(
  "Note: The `greed` package was not available when this vignette was\n",
  "built. The code chunks above were not evaluated. Install `greed`\n",
  "with install.packages('greed') to reproduce the examples.\n",
  sep = ""
)
```
