---
title: "Out-of-core RR computation with DuckDB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{monecascale-duckdb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5,
  dpi = 150,
  out.width = "100%"
)

# Eval-gate on duckdb (Suggests). If missing, skip evaluation and still
# render a readable article.
eval_duckdb <- requireNamespace("duckdb", quietly = TRUE) &&
  requireNamespace("DBI", quietly = TRUE)
knitr::opts_chunk$set(eval = eval_duckdb)

have_greed  <- requireNamespace("greed", quietly = TRUE)
have_igraph <- requireNamespace("igraph", quietly = TRUE)
```

## 1. Introduction: the RR bottleneck

MONECA clusters mobility data on the relative-risk matrix
`RR = O / E`, where `E_ij = r_i * c_j / T` is the expected count
under row / column independence. The reference implementation
`moneca::weight.matrix()` densifies the input on its way to RR
(`analytical_functions.R:581-595`): the observed count matrix `O` is
materialised in full before the ratio is formed. That densification
sets a hard R-side ceiling on problem size. A mobility graph on
10^5+ nodes may fit comfortably as a sparse edge list, yet its dense
matrix representation does not fit in R memory at all.

`rr_from_duckdb()` lifts the RR construction step out of R and into
DuckDB SQL. The margins, the grand total, and the cell-wise ratio
are computed as an aggregation over the edge list; only the
non-zero cells are ever materialised, and only once, on their way
into a sparse `dgCMatrix`. The densification is removed at its root
rather than worked around.

The output composes directly with monecascale's sparse-aware
backends. `moneca_bipartite()` consumes the RR matrix via its
rectangular argument path; `moneca_sbm()` and `moneca_flow()`
consume the raw counts exposed on `$counts`. The end result is that
mobility datasets an order of magnitude larger than
`moneca::moneca_fast()` can handle become tractable, without giving
up the moneca-class contract that downstream analysis and plotting
APIs rely on.

## 2. The SQL recipe

The RR computation is a three-CTE aggregation over the edge list.
One CTE sums the counts per row; one sums per column; one takes the
grand total. A final join multiplies the observed count by
`total / (row_sum * col_sum)` to yield `rr`.

```sql
WITH row_sums AS (
  SELECT row_id, SUM(w) AS s FROM edges GROUP BY row_id
),
col_sums AS (
  SELECT col_id, SUM(w) AS s FROM edges GROUP BY col_id
),
totals AS (
  SELECT SUM(w) AS total FROM edges
)
SELECT
  e.row_id,
  e.col_id,
  e.w AS count,
  e.w * t.total / (r.s * c.s) AS rr
FROM edges e
JOIN row_sums r ON r.row_id = e.row_id
JOIN col_sums c ON c.col_id = e.col_id
CROSS JOIN totals t;
```

No ordering is required, and no dense intermediate is constructed.
DuckDB streams the aggregations over the edge list, then streams
the join result back to R in long form. The R side receives a
`(row_id, col_id, count, rr)` table whose length equals the number
of non-zero cells in `O`, and feeds that directly into
`Matrix::sparseMatrix()`. `$margins$total` reflects the full input
total rather than a post-filter total, because that is what the RR
denominator uses.

## 3. Three input modes

`rr_from_duckdb()` accepts three shapes of input. The examples
below run on a small square fixture so they evaluate in a vignette
build; the same code shape scales to arbitrarily large edge lists
on disk.

### 3.1 In-memory data frame

```{r inmem}
library(monecascale)

set.seed(2026)
edges <- data.frame(
  row    = sample(paste0("n", 1:30), 200, replace = TRUE),
  col    = sample(paste0("n", 1:30), 200, replace = TRUE),
  weight = rpois(200, 3) + 1L,
  stringsAsFactors = FALSE
)

res <- rr_from_duckdb(edges, weight_col = "weight", verbose = TRUE)
str(res, max.level = 1)
```

The return value is a `"monecascale_rr"` list: `$rr` is the sparse
RR matrix, `$counts` the sparse observed count matrix,
`$margins$row / $col / $total` the raw margins (with `$total` being
the full pre-diagonal-filter grand total used as the RR
denominator), `$node_names$row / $col` the dimnames, `$bipartite`
the mode flag, and `$n_edges / $n_nonzero` integer summaries.

### 3.2 On-disk CSV or Parquet

When the edge list is already on disk, pass a path instead of a
data frame. DuckDB reads CSV natively via `read_csv_auto()`; R never
holds the edges in memory.

```{r ondisk-csv}
path <- tempfile(fileext = ".csv")
utils::write.csv(edges, path, row.names = FALSE)

res_csv <- rr_from_duckdb(path, weight_col = "weight")
identical(dim(res_csv$rr), dim(res$rr))
```

Parquet input works identically: DuckDB dispatches on the file
extension to `read_parquet()`. The Parquet path is not shown here
because it would introduce an `arrow` dependency for a
demonstration, but the calling convention is unchanged.

### 3.3 Pre-connected DuckDB table

The third mode is for users who already have a DuckDB connection
with data registered. When `edges` is a character scalar that does
not resolve to a file on disk, `rr_from_duckdb()` treats it as a
table or view name on the supplied connection. This is the pattern
for streaming Parquet datasets or remote tables that a user
attaches once and queries many times.

```{r preconn}
con <- DBI::dbConnect(duckdb::duckdb(), dbdir = ":memory:")
DBI::dbWriteTable(con, "edges_tbl", edges)

res_con <- rr_from_duckdb("edges_tbl", con = con, weight_col = "weight")
identical(dim(res_con$rr), dim(res$rr))

DBI::dbDisconnect(con, shutdown = TRUE)
```

## 4. Composition with `moneca_bipartite()`

For rectangular input (`bipartite = TRUE`) the RR matrix is the
natural input to `moneca_bipartite()`. A 30 x 20 fixture is small
enough to fit in a vignette and large enough to illustrate the
composition.

```{r bip-rr}
set.seed(2026)
edges_bip <- data.frame(
  row    = paste0("r", sample(1:30, 300, replace = TRUE)),
  col    = paste0("c", sample(1:20, 300, replace = TRUE)),
  weight = rpois(300, 3) + 1L
)

res_bip <- rr_from_duckdb(edges_bip, weight_col = "weight", bipartite = TRUE)
dim(res_bip$rr)
res_bip$bipartite
```

```{r bip-fit, eval = eval_duckdb && have_greed}
fit <- monecascale::moneca_bipartite(as.matrix(res_bip$rr))
print(fit)
```

The dense coercion `as.matrix(res_bip$rr)` is used here because the
current `moneca_bipartite()` contract is tested on dense input. At
true 10^5-row scale the dense cast defeats the purpose of the
out-of-core build; that path is reserved for a sparse-aware
bipartite fit planned as a follow-up. The point of this section is
the composition pattern, not the scale.

## 5. Composition with `moneca_sbm()` and `moneca_flow()` via `$counts`

`moneca_sbm()` and `moneca_flow()` consume raw mobility counts, not
RR: the SBM backend builds its own independence null internally,
and Infomap needs the unnormalised flow magnitudes to weight the
random walk. For those entry points the useful slot is `$counts`,
exposed alongside `$rr` specifically so that one DuckDB pass feeds
both downstream paths.

```{r counts}
res <- rr_from_duckdb(edges, weight_col = "weight")
dense_counts <- as.matrix(res$counts)
dim(dense_counts)
```

```{r flow-fit, eval = eval_duckdb && have_igraph}
fit_flow <- monecascale::moneca_flow(
  dense_counts,
  segment.levels = 2L,
  seed = 2026
)
print(fit_flow)
```

```{r sbm-fit, eval = eval_duckdb && have_greed}
fit_sbm <- monecascale::moneca_sbm(dense_counts, backend = "greed")
print(fit_sbm)
```

`$rr` is the input for paths that use the relative-risk matrix
directly, such as `moneca_bipartite()` and the planned
`moneca_localclique()`. `$counts` is the input for paths that
expect raw mobility. A single DuckDB aggregation yields both.

## 6. Limitations

- **No Arrow dataset input yet.** Partitioned Parquet datasets must
  be materialised through DuckDB's own `read_parquet()` rather than
  registered as Arrow datasets. A direct Arrow path is planned as a
  follow-up.
- **No streaming RR persistence.** The sparse RR is returned to R
  in memory. Users whose RR does not fit in R memory need to
  persist the long-format cell table through DuckDB directly; a
  streaming writer is not exposed.
- **Downstream ceiling from moneca metadata.**
  `moneca::moneca_segments()`, called internally by the SBM and
  flow adapters, still densifies `mat.list[[1]]` when it builds
  per-segment summaries. The full pipeline therefore retains an
  R-side ceiling around 10^5 nodes until upstream moneca adds a
  sparse metadata path. `rr_from_duckdb()` removes the RR-build
  bottleneck; it does not remove the metadata-build one.
- **Optional dependency.** `duckdb` is a `Suggests`. Install with
  `install.packages("duckdb")`. `rr_from_duckdb()` errors
  informatively when the package is not available.

## 7. References

- DuckDB: <https://duckdb.org/>.
- Raasveldt, M., & Muhleisen, H. (2019). DuckDB: an embeddable
  analytical database. *Proceedings of the 2019 International
  Conference on Management of Data*, 1981-1984.
- The monecascale scaling roadmap places `rr_from_duckdb()` under
  direction D7 (out-of-core / streaming backend), shipped in
  version 0.5.0 alongside the SBM, bipartite, flow, and auto-level
  backends documented in the other vignettes.

```{r no-duckdb-note, echo = FALSE, eval = !eval_duckdb}
cat(
  "Note: the `duckdb` package was not available when this vignette\n",
  "was built. The code chunks above were not evaluated. Install\n",
  "duckdb with install.packages('duckdb') to reproduce the examples.\n",
  sep = ""
)
```
