---
title: "monecascale: hierarchical DC-SBM backend for MONECA"
author: "Giampaolo Montaletti"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{monecascale: hierarchical DC-SBM backend for MONECA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5,
  dpi = 150,
  out.width = "100%"
)

# Eval-gate on greed (Suggests). If missing, skip evaluation and still
# render a readable article.
have_greed <- requireNamespace("greed", quietly = TRUE)
if (!have_greed) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

## About monecascale

`monecascale` is a sibling package to [`moneca`](https://github.com/gmontaletti/MONECA).
It provides **scalable clustering backends** whose output is consumed
unchanged by moneca's analysis and plotting stack. The first backend,
`moneca_sbm()`, replaces moneca's clique-enumeration step with
hierarchical degree-corrected stochastic block model (DC-SBM)
inference — the natural statistical twin of moneca's relative-risk
framing.

This vignette motivates the design, shows a minimal end-to-end
workflow, reproduces the empirical parity measurement against
`moneca::moneca_fast()`, and explains when to reach for which backend.

## The theoretical bridge: RR = O/E is the DC-SBM residual

MONECA's weight matrix is the relative-risk deviation of observed
mobility from the independence null:

$$
RR_{ij} = \frac{O_{ij}}{\bigl(r_i \cdot c_j\bigr) / T}.
$$

The degree-corrected stochastic block model (Karrer & Newman, 2011)
models the expected number of edges between nodes `i` and `j` as
`\theta_i \theta_j \omega_{g_i, g_j}`, with `\theta` degree parameters
and `\omega` a block-interaction matrix. Setting `\omega` to the
identity yields exactly `(r_i c_j) / T` as the expected count — i.e.,
MONECA's independence null. In that sense **MONECA's RR is the
DC-SBM residual**, and the clique rule on the thresholded RR is a
hand-rolled approximation to block inference. Any modern DC-SBM
inference routine is therefore a principled scalable drop-in.

Two backends are wired up behind `moneca_sbm()`:

- `"greed"` (default, R-native via `greed::DcSbm`): hierarchical
  dendrogram, Poisson likelihood for counts, sparse `dgCMatrix` input
  end-to-end, portable parallel via the `future` framework. Scales
  comfortably to ~10^4 – 10^5 nodes.
- `"graphtool"` (opt-in, Python via `reticulate`): Peixoto's
  `minimize_nested_blockmodel_dl()`, which uses merge-split MCMC to
  minimize description length directly. Scales to 10^6+ nodes on
  commodity hardware.

## Quick start

```{r quick-start}
library(monecascale)
library(moneca)

mob <- generate_mobility_data(
  n_classes = 20,
  n_total = 5000,
  immobility_strength = 0.5,
  class_clustering = 0.75,
  noise_level = 0.05,
  seed = 2026
)

seg <- moneca_sbm(mob, backend = "greed", seed = 2026)
seg
```

`moneca::segment.membership.dataframe()` provides the same tidy membership
view as for `moneca_fast()` output:

```{r membership}
head(segment.membership.dataframe(seg), 10)
```

## Hierarchy without choosing `segment.levels`

Unlike `moneca::moneca_fast()`, SBM backends return a full nested
hierarchy without requiring the user to pre-specify `segment.levels`.
The `$sbm_diagnostics` slot records the number of blocks and the
MDL-analogue score at every level:

```{r diagnostics}
seg$sbm_diagnostics$n_blocks_per_level
round(seg$sbm_diagnostics$mdl_per_level, 2)
```

The adapter preserves moneca's convention: `segment.list[[1]]` is
atomic and `segment.list[[l]]` for `l >= 2` runs from finest to
coarsest. Passing `segment.levels = 3` truncates the hierarchy to
three approximately log-spaced levels while preserving the finest and
coarsest.

```{r truncate}
seg3 <- moneca_sbm(mob, backend = "greed", seed = 2026, segment.levels = 3)
length(seg3$segment.list)
lengths(seg3$segment.list)
```

## Empirical parity against `moneca::moneca_fast()`

If the theoretical bridge is right, `moneca_sbm(backend = "greed")`
should recover segmentations close to those of `moneca::moneca_fast()`
on a well-structured mobility matrix. We measure this directly with the
adjusted Rand index (ARI) and normalized mutual information (NMI)
between the per-node labelings produced by both methods.

```{r parity-helpers}
ari <- function(a, b) {
  tab <- table(a, b); n <- sum(tab)
  sc <- function(x) sum(choose(x, 2))
  a_sum <- sc(rowSums(tab)); b_sum <- sc(colSums(tab))
  tab_sum <- sc(as.vector(tab))
  expected <- a_sum * b_sum / choose(n, 2)
  max_idx <- (a_sum + b_sum) / 2
  if (max_idx == expected) return(1)
  (tab_sum - expected) / (max_idx - expected)
}

nmi <- function(a, b) {
  tab <- table(a, b); n <- sum(tab); p <- tab / n
  pa <- rowSums(p); pb <- colSums(p); nz <- p > 0
  mi <- sum(p[nz] * log(p[nz] / outer(pa, pb)[nz]))
  ha <- -sum(pa[pa > 0] * log(pa[pa > 0]))
  hb <- -sum(pb[pb > 0] * log(pb[pb > 0]))
  if (ha == 0 || hb == 0) return(0)
  mi / sqrt(ha * hb)
}

level_labels <- function(mon, level) {
  n_core <- nrow(mon$mat.list[[1]]) - 1L
  labels <- rep(NA_integer_, n_core)
  cliques <- mon$segment.list[[level]]
  for (b in seq_along(cliques)) labels[cliques[[b]]] <- b
  na_idx <- is.na(labels)
  if (any(na_idx)) {
    labels[na_idx] <- max(0L, labels, na.rm = TRUE) + seq_len(sum(na_idx))
  }
  as.integer(labels)
}
```

```{r parity}
fast <- moneca::moneca_fast(mob, segment.levels = 3, progress = FALSE)
sbm  <- moneca_sbm(mob, backend = "greed", seed = 2026)

fast_levels <- seq.int(2L, length(fast$segment.list))
sbm_levels  <- seq.int(2L, length(sbm$segment.list))

grid <- expand.grid(fast_lvl = fast_levels, sbm_lvl = sbm_levels)
grid$ari <- mapply(function(lf, ls) ari(level_labels(fast, lf),
                                        level_labels(sbm, ls)),
                   grid$fast_lvl, grid$sbm_lvl)
grid$nmi <- mapply(function(lf, ls) nmi(level_labels(fast, lf),
                                        level_labels(sbm, ls)),
                   grid$fast_lvl, grid$sbm_lvl)

best <- grid[which.max(grid$ari), , drop = FALSE]
best
```

On this 20-class synthetic fixture the best matching level pair
typically recovers `ARI ~ 0.67` and `NMI ~ 0.90` — above the
`ARI >= 0.55` threshold set as the aspirational target when the
direction was planned, and strong empirical confirmation of the
RR ≡ DC-SBM bridge.

The match is not exact — and should not be. Clique enumeration and
block-model inference are structurally different procedures: cliques
insist on internal completeness at a threshold, blocks allow any
stochastically-consistent edge pattern within a group. When the two
agree on segmentation, the mobility data has sharply modular block
structure; when they disagree, the discrepancy is itself diagnostic.

## Choosing a backend

| Criterion                       | `backend = "greed"`             | `backend = "graphtool"`                          |
|---------------------------------|---------------------------------|--------------------------------------------------|
| Runtime install cost            | `install.packages("greed")`     | conda env + `graph-tool` (~800 MB)               |
| Scale ceiling (nodes)           | ~10^4 – 10^5                    | ~10^6 – 10^7                                     |
| Sparse `dgCMatrix` input        | yes, end-to-end                 | via reticulate edge list                         |
| Hierarchy construction          | merge dendrogram; `cut(sol,k)`  | nested MCMC, picks depth by MDL                  |
| Determinism under `seed`        | high (deterministic local search)    | MCMC; use `n_init` best-of-N plus `seed`    |
| Parallel                        | `future` framework, portable    | OpenMP; single-threaded recommended for `seed`   |
| Portability                     | pure R                          | Python + conda                                   |

Rule of thumb: use `greed` until you exceed roughly 5000 nodes, then
move to `graphtool`. The `backend = "auto"` default encodes this rule
and falls back to `greed` with an informational message if `graph-tool`
is not installed.

## Installing the `graph-tool` backend

`graph-tool` is not distributed on PyPI, and the macOS homebrew formula
compiles from source and breaks on Boost/Python version bumps. The
supported route is conda-forge. `moneca_sbm_install_graphtool()` wraps
the conda step:

```r
monecascale::moneca_sbm_install_graphtool(envname = "moneca-sbm")
reticulate::use_condaenv("moneca-sbm", required = TRUE)
```

After this, `moneca_sbm(backend = "graphtool")` works identically to
`"greed"`. A typical install takes 2–5 minutes on a warm conda cache.

## Downstream compatibility

Every moneca function that consumes a `moneca` object consumes a
`monecascale::moneca_sbm()` output unchanged:

```{r downstream-membership}
moneca::segment.membership(seg)[1:6, ]
```

```{r downstream-plot-hier, fig.alt = "Hierarchical plot of SBM segmentation"}
if (requireNamespace("ggraph", quietly = TRUE)) {
  moneca::plot_moneca_hierarchical(seg, title = "monecascale - greed backend")
}
```

```{r downstream-plot-ggraph, fig.alt = "Network plot of finest SBM level"}
if (requireNamespace("ggraph", quietly = TRUE)) {
  moneca::plot_moneca_ggraph(
    seg, level = 2, node_color = "segment",
    title = "monecascale - finest level"
  )
}
```

```{r downstream-dendro, fig.alt = "Dendrogram of SBM hierarchy", fig.height = 7}
if (requireNamespace("ggraph", quietly = TRUE)) {
  moneca::plot_moneca_dendrogram(seg)
}
```

## Limitations and future directions

`moneca_sbm()` is the first of seven directions in the MONECA scaling
roadmap. Known limits on this release:

- **Rectangular input is not accepted.** Person × employer matrices are
  the target use case for direction **D1** (`moneca_bipartite()`);
  `moneca_sbm()` currently errors on non-square input with a pointer
  there.
- **Auto-level unification** across backends (`segment.levels = "auto"`
  everywhere including `moneca::moneca_fast()`) is direction **D6**.
- **Out-of-core inference** for data that does not fit in RAM is
  direction **D7** (`moneca_duckdb` / `moneca_arrow`).

## References

- Karrer, B., & Newman, M. E. J. (2011). Stochastic blockmodels and
  community structure in networks. *Physical Review E*, 83, 016107.
- Peixoto, T. P. (2014). Hierarchical block structures and
  high-resolution model selection in large networks. *Physical Review X*,
  4, 011047.
- Côme, E., & Jouvin, N. (2020). `greed`: Clustering and Model Selection
  with the Integrated Classification Likelihood. R package.
- Touboel, J., & Larsen, A. G. (2017). Mapping the Social Class
  Structure: From Occupational Mobility to Social Class Categories
  Using Network Analysis. *Sociology*, 51(6), 1257–1276.

```{r no-greed-note, echo = FALSE, eval = !have_greed}
cat(
  "Note: The `greed` package was not available when this vignette was built.\n",
  "The code chunks above were not evaluated. Install `greed` with\n",
  "install.packages('greed') to reproduce the parity experiment.\n",
  sep = ""
)
```
