---
title: "Flow Clustering with moneca_flow()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{monecascale-flow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5,
  dpi = 150,
  out.width = "100%"
)

# Eval-gate on igraph (the only flow backend shipped in 0.4.0) and on
# moneca (always present as a hard dependency, but gating keeps the
# vignette renderable outside an installed environment).
have_igraph <- requireNamespace("igraph", quietly = TRUE)
have_moneca <- requireNamespace("moneca", quietly = TRUE)
if (!(have_igraph && have_moneca)) {
  knitr::opts_chunk$set(eval = FALSE)
}

have_greed   <- requireNamespace("greed", quietly = TRUE)
have_mclust  <- requireNamespace("mclust", quietly = TRUE)
have_ggraph  <- requireNamespace("ggraph", quietly = TRUE)
have_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## 1. Introduction

`moneca_flow()` is the flow-based clustering backend for MONECA. It
clusters the mobility graph under the Map Equation (Rosvall &
Bergstrom, 2008): a random walker traverses the directed mobility
network and modules are extracted as regions where the walker tends
to stay for many steps before transitioning away. This reading aligns
directly with the substantive semantics of the data — workers flow
through jobs, and modules trap flow.

Three backends cover complementary views of the same mobility matrix:

- `moneca::moneca_fast()` — clique enumeration on the thresholded
  relative-risk (RR) matrix.
- `moneca_sbm()` (direction D2) — hierarchical DC-SBM on the RR
  residual. See `vignette("monecascale-sbm")` for the theoretical
  bridge and an empirical parity experiment.
- `moneca_bipartite()` (direction D1) — hierarchical DC-LBM on
  rectangular mobility, joint on both sides.
- `moneca_flow()` (direction D3, this vignette) — Infomap on raw
  mobility flow, with a recursive wrapper that synthesises an
  approximate hierarchy.

The SBM view clusters nodes that are stochastically equivalent under
the independence null; the flow view clusters nodes that the walker
visits together. The two views frequently agree in the high-signal
regime and may diverge where directional asymmetry matters. Section 7
shows a side-by-side comparison on a single fixture.

## 2. The Map Equation

The Map Equation frames community detection as a lossless coding
problem. Given a partition `M` of the graph into modules, the
walker's trajectory is encoded with a two-level codebook: an
index-level codebook for module-to-module transitions, and one
per-module codebook for within-module node visits plus a module exit
symbol. The average number of bits per step under this encoding is
the codelength

$$
L(M) = q_\curvearrowleft\, H(\mathcal{Q}) + \sum_{k=1}^{K} p_k^{\circlearrowright}\, H(\mathcal{P}_k),
$$

where `q_↶` is the total module-switching rate, `H(Q)` is the entropy
of the index codebook, `p_k^↻` is the within-module use rate of
module `k`, and `H(P_k)` is the entropy of its within-module
codebook. Minimising `L(M)` is equivalent to maximising the
compression of the walker's trajectory: modules that trap flow for
many steps shorten the description, modules that shuffle flow around
lengthen it.

In MONECA the edges are directed mobility counts, so the stationary
distribution used to weight the entropies is computed from PageRank
on the directed graph. `moneca_flow()` computes `L(M)` directly on
the full graph for every level of the hierarchy, so codelength values
are comparable across levels rather than inferred from sub-module
fits.

## 3. Flat vs recursive hierarchy

`igraph::cluster_infomap()`, the engine behind the `"igraph"`
backend, returns a flat partition. To expose a hierarchy compatible
with moneca's multi-level contract, `moneca_flow()` applies a
recursive wrapper: for level 2 it re-runs flat Infomap on each
level-1 module's induced subgraph; for level 3 it recurses once more
on each level-2 sub-module. Parent modules smaller than two nodes
become singleton children.

This is an **approximation** to true hierarchical Infomap. The
orthodox hierarchical Map Equation (Rosvall et al., 2009) jointly
optimises the codelength of the full hierarchy, trading off top-level
compression against sub-module compression in a single criterion.
Recursive flat fits instead optimise each level independently and
then stack the results. On well-separated block structures the two
agree closely; on ambiguous or deeply nested hierarchies the
recursive approximation can under- or over-fit specific levels.

Users who need joint-MDL hierarchical Infomap should watch for a
`bioregion`-backed or `"infomap_cpp"` backend planned for a future
release. The recursive flat approach is shipped now because it keeps
the dependency surface to `igraph` alone.

## 4. Quick start

A minimal end-to-end run on a synthetic block-Poisson fixture:

```{r quick-start}
library(monecascale)

# 1. inline block-Poisson fixture -----
set.seed(2026)
n <- 21
block <- rep(1:3, each = 7)
lambda <- outer(block, block, function(g, h) ifelse(g == h, 6, 0.2))
M <- matrix(stats::rpois(n * n, lambda = lambda), n, n)
diag(M) <- 0
rownames(M) <- colnames(M) <- paste0("N", seq_len(n))

# 2. flow fit -----
out <- monecascale::moneca_flow(M, depth = 2L, seed = 2026)
out

# 3. codelength trace and per-level block counts -----
out$flow_diagnostics$codelength_per_level
out$flow_diagnostics$n_blocks_per_level
```

`out` is a regular moneca-class object: `segment.list`, `mat.list`,
and `segment_metadata` follow the same contract as
`moneca::moneca_fast()`. The `$flow_diagnostics` slot adds the flow-
specific trace (codelength per level, block counts, backend, depth).
`$flow_diagnostics$mdl_per_level` is an alias of
`$codelength_per_level` so tooling keyed on `$mdl_per_level` consumes
flow output unchanged.

## 5. Auto-level selection

The auto-level machinery (`auto_segment_levels()`, direction D6)
composes with flow fits without modification. The MDL criterion
reads the codelength trace under the hood; diagnostics are exposed
in a `codelength` column rather than `mdl` to match the flow
semantics.

```{r auto-post-hoc}
# 1. post-hoc pick from a full-hierarchy fit -----
pick <- monecascale::auto_segment_levels(out, method = "mdl")
pick
pick$diagnostics
```

The wrapper form trims the fit to the picked level in one call:

```{r auto-wrapper}
out_auto <- monecascale::moneca_flow(
  M,
  depth = 2L,
  seed = 2026,
  segment.levels = "auto",
  auto_method = "mdl"
)

# Number of levels retained after trimming, plus the picker record
length(out_auto$segment.list)
out_auto$auto_level$level
```

`method = "mi_plateau"` works identically on flow fits: it reads
memberships from `segment.list` and never touches the backend trace,
so it is backend-agnostic.

## 6. Downstream compatibility

Flow fits reuse moneca's entire analysis and plotting stack.
Membership extraction and segment quality diagnostics run unchanged:

```{r downstream-membership}
# 1. level-2 memberships as a vector -----
mem_l2 <- moneca::segment.membership(out, level = 2)
head(mem_l2)
```

```{r downstream-quality}
# 2. moneca's segment-quality summary on the flow fit -----
moneca::segment.quality(out)
```

ggraph-based plotting is gated on the optional `ggraph` dependency:

```{r downstream-ggraph, eval = have_igraph && have_moneca && have_ggraph}
moneca::plot_moneca_ggraph(out, level = 2)
```

## 7. SBM vs Flow on the same fixture

The SBM and flow views answer different structural questions on the
same data. Running both on a single fixture makes the comparison
concrete. We reuse the fixture from section 4 and fit both backends
with two recursion / hierarchy levels.

```{r compare-fits, eval = have_igraph && have_moneca && have_greed}
# 1. flow fit -----
fit_flow <- monecascale::moneca_flow(M, depth = 2L, seed = 2026)

# 2. SBM fit (greed backend) -----
fit_sbm <- monecascale::moneca_sbm(
  M,
  backend = "greed",
  seed = 2026
)

# 3. level-1 memberships from each -----
mem_flow <- moneca::segment.membership(fit_flow, level = 2)
mem_sbm  <- moneca::segment.membership(fit_sbm,  level = 2)

length(unique(mem_flow))
length(unique(mem_sbm))
```

If `mclust` is available we can quantify agreement with the adjusted
Rand index and normalised mutual information.

```{r compare-ari, eval = have_igraph && have_moneca && have_greed && have_mclust}
ari <- mclust::adjustedRandIndex(mem_flow, mem_sbm)
nmi <- {
  # 1. plug-in NMI from a contingency table -----
  tab <- table(mem_flow, mem_sbm)
  p   <- tab / sum(tab)
  pr  <- rowSums(p)
  pc  <- colSums(p)
  H   <- function(x) {
    x <- x[x > 0]
    -sum(x * log(x))
  }
  mi <- sum(p[p > 0] * log(p[p > 0] / outer(pr, pc)[p > 0]))
  mi / sqrt(H(pr) * H(pc))
}
c(ARI = ari, NMI = nmi)
```

On block-Poisson fixtures with clean planted structure the two
backends converge: flow and SBM recover the same partition because
both the stationary flow and the stochastic-equivalence signal point
at the same blocks. They diverge when directional asymmetry carries
real structure — for instance, when a transient module feeds flow
into an absorbing one: flow tends to merge them into a single trap,
SBM keeps them distinct because their in/out degree profiles differ.
Running both and inspecting disagreement is a practical diagnostic.

## 8. Limitations

- **Recursive flat approximates joint-MDL hierarchical Infomap.** The
  shipped wrapper optimises each level independently. For strict
  hierarchical codelength minimisation use the standalone Infomap
  C++ binary; a `bioregion`-backed backend is planned for v0.5.0.
- **No overlapping communities.** The underlying Map Equation
  supports overlapping modules via the two-level memory-based
  variant, but `moneca_flow()` produces hard partitions only. The
  moneca contract assumes hard membership vectors.
- **Raw mobility counts only.** `moneca_flow()` does not apply the
  RR transformation. Users who want RR-weighted flow clustering
  should preprocess the matrix themselves, or reach for
  `moneca_sbm()` which consumes RR natively via the DC-SBM bridge.
- **Single backend in 0.4.0.** Only the `igraph` backend is shipped.
  The `"infomap_cpp"` keyword is reserved but not wired up.
- **Depth capped at 3.** The recursive wrapper supports `depth` in
  `1:3`. Deeper hierarchies rarely add signal and are better served
  by auto-level selection on a full SBM hierarchy.

## 9. References

- Rosvall, M., & Bergstrom, C. T. (2008). Maps of random walks on
  complex networks reveal community structure. *PNAS*, 105(4),
  1118-1123.
- Rosvall, M., Axelsson, D., & Bergstrom, C. T. (2009). The map
  equation. *European Physical Journal Special Topics*, 178,
  13-23.
- Satopaa, V., Albrecht, J., Irwin, D., & Raghavan, B. (2011).
  Finding a "Kneedle" in a Haystack: Detecting Knee Points in
  System Behavior. *31st International Conference on Distributed
  Computing Systems Workshops*, 166-171.
- Karrer, B., & Newman, M. E. J. (2011). Stochastic blockmodels and
  community structure in networks. *Physical Review E*, 83, 016107.

```{r no-igraph-note, echo = FALSE, eval = !(have_igraph && have_moneca)}
cat(
  "Note: igraph (and/or moneca) was not available when this vignette\n",
  "was built. The code chunks above were not evaluated. Install\n",
  "igraph with install.packages('igraph') to reproduce the examples.\n",
  sep = ""
)
```
